---
title: "Final Project EDA 4.17"
author: "West Coasters"
format: html
---

# Set Up

```{r load_packages, warning=FALSE, message=FALSE}
# R packages
library(ipumsr)
library(tidyverse)
library(dplyr)
library(ggplot2)
```

```{r load_data, warning=FALSE, message=FALSE}
ddi <- read_ipums_ddi("data/ahtus_00001.xml")
data <- read_ipums_micro(ddi)
```


# Questions: Exploring the Data

## What is your outcome variable(s)? How well does it measure the outcome you are interested in? How does it relate to your expectations?

  * Considered Wage
  * Paid Work
  * Considered Personal Care (nothing significant) 

We did several multiple linear regressions with wage as the outcome variable. We expected wages to have significant results, when explanatory variables such as gender, childcare, and an interaction term of $(gender \times childcare)$ were included. We expected there to be a larger wage difference from the $(gender \times childcare)$ interaction term. However, we noticed small and insignificant coefficients for this regression. We’d expect time spent on childcare to be greater for women than men. 

Further, there is a tradeoff between time spent at paid work versus time spent on childcare. So, to interrogate this idea of the time tradeoff, we ran a subsequent regression with time spent on paid work as the outcome variable. This regression includes gender and childcare as explanatory variables, with the $(gender \times childcare)$ interaction term. Here, we found significant results and a clear visual relationship of more time spent on childcare meant less time spent on paid work. This upheld our expectations, as both women and men exhibited this relationship. 


## What are your key explanatory variables? 

  * Gender binary
  * Time spent on childcare
  * Time spent on unpaid domestic work
  * Time spent on paid work
  * Education: 1 if EDUC == 5 or 6 (college graduate), 0 if 1 through 4


## Research Questions

Is there a gender difference in how individuals allocate their time? How does the tradeoff between time spent at paid work versus time spent on childcare differ by gender? Ultimately, the question of interest is how do demographics and expected responsibilities affect time use.



# Questions: Data Wrangling and Transformation

## What data cleaning did you do?

  * Converted the time variables from minutes to hours for easy interpretation
  * Removed rows with NAs
  * Other than that, the dataset was already well cleaned by IPUMs! 🙂

## How did you wrangle the data?

Applied the following filters:
  * Married with kids: HHTYPE == 1
  * Child under 18: UNDER18 1 through 10
  * Sample: 2010-2018
  * Post-college education: EDUC == 5 or 6
  * Full-time workers: FULLTIME == 1 (Only comparing men and women who work full-time)

Sometimes this varied by different figures for less or more granular visualizations.

## Are you deciding to exclude any observations? If so, why?

Yes, but this was mostly the data wrangling step. We removed any missing values or NAs. For some visualizations (Figure 2) a few of the time use variables were omitted to better depict explanatory variables, rather than activities we’re less interested in: such as travel, media consumption, physical activity, etc.

Did you have to create any new variables from existing variables? If so, how and why?
Yes, when performing PCA on the time variables, we decided to combine a few variables to increase interpretability.
ACT_INHOME and ACT_OUTHOME, which both represented leisure, were combined into ACT_LEISURE.
ACT_PCARE and ACT_PHYSICAL, which represented personal care and physical activity, were combined into ACT_SELFCARE since physical activity (exercise) is also a form of personal/self care.


# Data Cleaning

## Filtering for PCA Visualization 

```{r}
# data with filters needed for PCA visualizations (Figures 3 and 4)
filtered_data <- data %>%
  filter(
    CIVSTAT == 1,
    HHTYPE == 1,
    UNDER18 >= 1,
    AGE >= 22 & AGE <= 40,
    EDUC %in% c(5, 6),
    AGEYNGST >= 0 & AGEYNGST <= 18,
    SEX %in% c(1, 2)
  ) %>%
  mutate(
    sex_label = ifelse(SEX == 1, "Male", "Female")
  )
```

## Filtering for Regression Model

This is the primary data cleaning and transforming done. 

```{r clean_transform}

# filtering and turning sex and full-time into binary variables
datareg <- data %>%
  
  filter(
    HHTYPE == 1, # filter to married with kids
                 # isolating a group where childcare is relevant
                 # avoid comparing people who don’t face childcare constraints
    SEX %in% c(1, 2), # filter out missing or dirty cases of sex
    # FULLTIME %in% c(0, 1), # filter out missing or dirty cases of full time work
    UNDER18 > 0, # filter to individuals with kids under age of 18
    !is.na(AGE), 
    !is.na(EDUC),
    FULLTIME == 1, # filter to just full-time workers
    (EDUC == 5 | EDUC == 6), # filter to just college educated individuals
    WAGELM > 0, # remove non-earners
    ACT_CHCARE >= 0,
    ACT_PCARE >= 0
  ) %>% 
  
  mutate(
    across(starts_with("ACT_"), ~ .x / 60), # convert all time use vars to hours
    female = ifelse(SEX == 2, 1, 0),
    female = factor(female),
    FULLTIME = factor(FULLTIME), 
    log_wage = log(WAGELM + 1) # add column for log wage in case it becomes interesting
  )
  
  
# filter to just 2010s data
data2010s <- datareg %>% 
  filter(SAMPLE > 2009)

```



# Codebook

https://github.com/ccaitlingo/westcoasters-finalproject/blob/main/README.md



# Exploratory Data Analysis

## Motivation 

### Figure 1: Unpaid Domestic Work by Sex

```{r}
# Unpaid Domestic Work by Sex (1930-2018)

# Average hours by year and sex
summary_data <- filtered_data %>%
  group_by(YEAR, sex_label) %>%
  mutate(ACT_UNDOM_HRS = ACT_UNDOM / 60) %>%
  summarise(avg_domestic_hours = mean(ACT_UNDOM_HRS, na.rm = TRUE), .groups = "drop")

# Line plot
ggplot(summary_data, aes(x = YEAR, y = avg_domestic_hours, color = sex_label)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "Unpaid Domestic Work by Sex (1930 - 2018)",
    x = "Year",
    y = "Average Hours per Day",
    color = "Sex"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("Male" = "steelblue", "Female" = "salmon"))

# Note for 2018
# 2.72 hrs for female
# 1.57 hrs for male
```

### Figure 2: Average Daily Time Use by Sex

```{r}
# How time is split by gender on an average day!

# filter by 
df <- data %>%
  filter(SAMPLE >= 2010 ) %>%  
  mutate(HHTYPE = as.numeric(as.factor(HHTYPE))) %>% 
  filter(HHTYPE == 1) %>% 
  filter(UNDER18 > 0) %>% 
  mutate(
    Leisure = ACT_INHOME + ACT_OUTHOME
  ) %>% 
  select(-"ACT_CIVIC", -"ACT_EDUCA", -"ACT_MEDIA", -"ACT_MISSING",-"ACT_INHOME", -"ACT_OUTHOME", -"ACT_TRAVEL", -"ACT_PHYSICAL" )
# pivot! to have all activities listed under activity_category
df_long <- df %>%
  pivot_longer(
    cols = starts_with("ACT_") | starts_with("Leisure"),
    names_to = "activity_category",
    values_to = "minutes"
  )  %>% 
  mutate(
    # Clean up the activity names (optional)
    activity_category = gsub("ACT_", "", activity_category),
    activity_category = case_when(
      activity_category == "CHCARE" ~ "Childcare",
      activity_category == "PCARE"  ~ "Personal Care",
      activity_category == "UNDOM"  ~ "Unpaid Domestic Work",
      activity_category == "WORK"   ~ "Paid Work",
      TRUE ~ activity_category
    ),
    # Convert SEX codes to labels
    SEX = as.numeric(as.factor(SEX)), 
    SEX = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female"
    )
  )

# aggregate the activities!
df_sum <- df_long %>%
  group_by(SEX, activity_category) %>%
  summarize(minutes = mean(minutes, na.rm = TRUE))


#plot
ggplot(df_sum, aes(x = SEX, y = minutes, fill = activity_category)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(
    x = "Gender",
    y = "Proportion of Time Spent per Day",
    fill = "Activity",
    title = "Average Daily Time Use by Sex 2010-2018"
  ) +
  theme_minimal(base_size = 13) +
theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    legend.title = element_text(face = "bold")
  )
```


### Figure 3: Clustering in 2D

```{r}
# Redefine dataset
filtered_data <- data %>%
  filter(
    CIVSTAT == 1,
    HHTYPE %in% c(1,2),
    AGE >= 22 & AGE <= 40,
    EDUC %in% c(5, 6),
    AGEYNGST >= 0 & AGEYNGST <= 18,
    SEX %in% c(1, 2)
  ) %>%
  mutate(
    sex_label = ifelse(SEX == 1, "Male", "Female")
  )
```

```{r, warning=FALSE, message=FALSE}
# Employment mapped onto clusters

# Combine two leisure categories
filtered_data$ACT_LEISURE <- filtered_data$ACT_INHOME + filtered_data$ACT_OUTHOME
filtered_data$ACT_SELFCARE <- filtered_data$ACT_PCARE + filtered_data$ACT_PHYSICAL

# Select time-use variables
time_vars <- c(
  "ACT_WORK", "ACT_CHCARE", "ACT_EDUCA",
  "ACT_LEISURE", "ACT_MEDIA", "ACT_TRAVEL", "ACT_SELFCARE",
  "ACT_UNDOM", "ACT_CIVIC", "AGEYNGST", "UNDER5", "UNDER18"
)

# Keep only selected variables + employement status
df_time <- filtered_data[, c(time_vars, "SEX")]

# Drop missing values
df_time <- na.omit(df_time)

# Sample data (needed for clustering stability)
set.seed(123)
df_sample <- df_time[sample(nrow(df_time), 5000), ]

# Create label
df_sample$binary_var <- ifelse(df_sample$SEX == 1, "Male",
                    ifelse(df_sample$SEX == 2, "Female",
                           "Non-Binary"))

# Standardize time-use variables
df_scaled <- scale(df_sample[, time_vars])

# Run PCA
pca_res <- prcomp(df_scaled, center = TRUE, scale. = TRUE)

# Optional scree plot (to determine how many PCs to use)
eig_vals <- pca_res$sdev^2
var_explained <- eig_vals / sum(eig_vals)

plot(var_explained, type = "b",
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     main = "Scree Plot (PCA)",
     pch = 16)

# View the principal components
pca_res$rotation[, 1:4]

# Extract first 4 PCs
pca_scores <- as.data.frame(pca_res$x[, 1:4])

# Hierarchical clustering on PCA space
dist_mat <- dist(pca_scores)
hc <- hclust(dist_mat, method = "ward.D2")

# Cut into clusters
k <- 4
clusters <- cutree(hc, k = k)

# Attach cluster + employement status
pca_scores$cluster <- factor(clusters)
pca_scores$binary_var <- df_sample$binary_var

# -------------------------
# Plot 1: Employement Status only
# -------------------------
plot(pca_scores$PC1, pca_scores$PC2,
     col = ifelse(pca_scores$binary_var == "Male", "blue", "red"),
     pch = 1,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "PCA Space Colored by Sex")

legend("topright",
       legend = c("Male", "Female"),
       col = c("blue", "red"),
       pch = 16,
       bty = "n")

# -------------------------
# Plot 2: Clusters + Binary Variable
# -------------------------
plot(pca_scores$PC1, pca_scores$PC2,
     col = pca_scores$cluster,
     pch = ifelse(pca_scores$binary_var == "Male", 1, 1),
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Clusters in PCA Space")

legend("topright",
       legend = c("Cluster 1","Cluster 2","Cluster 3","Cluster 4"),
       col = c(1:4, "black"),
       pch = 16,
       bty = "n")

# -------------------------
# Table
# -------------------------
# Counts
tab <- table(pca_scores$cluster, pca_scores$binary_var)

# Row-wise percentages (within cluster)
pct <- round(prop.table(tab, margin = 1) * 100, 1)
pct

# Group clusters 1+3 and 2+4
grp <- ifelse(pca_scores$cluster %in% c(1,3), "C1+3", "C2+4")
pct_grp <- round(prop.table(table(grp, pca_scores$binary_var), 1) * 100, 1)
pct_grp
```

### Figure 4: Clustering in 3D

```{r}
# Load packages
library(plotly)

# Combine two leisure categories
filtered_data$ACT_LEISURE <- filtered_data$ACT_INHOME + filtered_data$ACT_OUTHOME
filtered_data$ACT_SELFCARE <- filtered_data$ACT_PCARE + filtered_data$ACT_PHYSICAL

# Select time-use variables
time_vars <- c(
  "ACT_WORK", "ACT_CHCARE", "ACT_EDUCA",
  "ACT_LEISURE", "ACT_MEDIA", "ACT_TRAVEL", "ACT_SELFCARE",
  "ACT_UNDOM", "ACT_CIVIC", "AGEYNGST", "UNDER5", "UNDER18"
)

# Keep complete cases
df_time <- na.omit(filtered_data[, time_vars])

# Standardize data
df_scaled <- scale(df_time)

# Run PCA
pca_res <- prcomp(df_scaled, center = TRUE, scale. = TRUE)

# Extract first 4 principal components
scores <- as.data.frame(pca_res$x[, 1:4])

# Run clustering (optional, if you want colors)
set.seed(123)
clusters <- kmeans(scores, centers = 4, nstart = 25)$cluster
scores$cluster <- factor(clusters)

# 3D PCA plot
plot_ly(
  scores,
  x = ~PC1,
  y = ~PC2,
  z = ~PC3,
  color = ~cluster,
  type = "scatter3d",
  mode = "markers"
)
```


```{r}
# Load package
library(plotly)

# -------------------------
# Variables
# -------------------------
# Combine two leisure categories
filtered_data$ACT_LEISURE <- filtered_data$ACT_INHOME + filtered_data$ACT_OUTHOME
filtered_data$ACT_SELFCARE <- filtered_data$ACT_PCARE + filtered_data$ACT_PHYSICAL

# Select time-use variables
time_vars <- c(
  "ACT_WORK", "ACT_CHCARE", "ACT_EDUCA",
  "ACT_LEISURE", "ACT_MEDIA", "ACT_TRAVEL", "ACT_SELFCARE",
  "ACT_UNDOM", "ACT_CIVIC", "AGEYNGST", "UNDER5", "UNDER18"
)

# Keep data + sex, drop missing
df_time <- na.omit(filtered_data[, c(time_vars, "SEX")])

# -------------------------
# Gender label
# -------------------------
df_time$gender <- ifelse(df_time$SEX == 1, "Male",
                   ifelse(df_time$SEX == 2, "Female", "Non-Binary"))

# -------------------------
# Standardize
# -------------------------
df_scaled <- scale(df_time[, time_vars])

# -------------------------
# PCA (4 components)
# -------------------------
pca_res <- prcomp(df_scaled, center = TRUE, scale. = TRUE)

scores <- as.data.frame(pca_res$x[, 1:4])

# -------------------------
# Clustering (k-means)
# -------------------------
set.seed(123)
clusters <- kmeans(scores, centers = 4, nstart = 25)$cluster
scores$cluster <- factor(clusters)

# Add gender
scores$gender <- df_time$gender

# -------------------------
# 3D PCA plot (colored by sex)
# -------------------------
plot_ly(
  scores,
  x = ~PC1,
  y = ~PC2,
  z = ~PC3,
  color = ~gender,
  colors = c("blue", "red"),
  type = "scatter3d",
  mode = "markers"
) %>%
  layout(
    title = "3D PCA Space Colored by Sex",
    scene = list(
      xaxis = list(title = "PC1"),
      yaxis = list(title = "PC2"),
      zaxis = list(title = "PC3")
    )
  )

# -------------------------
# 3D PCA plot (colored by sex)
# -------------------------
# Counts
tab <- table(scores$cluster, scores$gender)

# Row-wise percentages (within each cluster)
pct <- round(prop.table(tab, margin = 1) * 100, 1)
pct

```


## Figure 5: Exploring the Research Question

```{r, warning=FALSE, message=FALSE}

data2010s %>%
  mutate(chcare_bin = ntile(ACT_CHCARE, 5)) %>%
  group_by(chcare_bin, female) %>%
  summarise(
    avg_wage = mean(ACT_WORK, na.rm = TRUE),
    avg_chcare = mean(ACT_CHCARE)
  ) %>%
  ggplot(aes(x = avg_chcare, y = avg_wage, color = female)) +
  geom_line() +
  geom_point() +
    labs(
      title = "Time Spent on Paid Work vs. Time Spent on Childcare by Gender",
      subtitle = "In the 2010s for Individuals who are College-Educated and Married with Kids Under 18",
      x = "Average Hours of Childcare per Day",
      y = "Average Hours of Paid Work per Day"
    ) 

```