---
title: "Final Project EDA 4.17"
author: "West Coasters"
format: html
editor: 
  markdown: 
    wrap: 72
---

# Set Up

```{r load_packages, warning=FALSE, message=FALSE}
# R packages
library(ipumsr)
library(tidyverse)
library(dplyr)
library(ggplot2)
```

```{r load_data, warning=FALSE, message=FALSE}
ddi <- read_ipums_ddi("data/ahtus_00001.xml")
data <- read_ipums_micro(ddi)
```

# Questions: Exploring the Data

## What is your outcome variable(s)? How well does it measure the outcome you are interested in? How does it relate to your expectations?

Our primary outcome variable is time spent on paid work. However, we
initially ran regressions with wage as the outcome variable and found
statistically insignificant results.

We did several multiple linear regressions with wage as the outcome
variable. We expected wages to have significant results, when
explanatory variables such as gender, childcare, and an interaction term
of $(gender \times childcare)$ were included. We expected there to be a
larger wage difference from the $(gender \times childcare)$ interaction
term. However, we noticed small and insignificant coefficients for this
regression. We’d expect time spent on childcare to be greater for women
than men.

Further, there is a trade-off between time spent at paid work versus
time spent on childcare. So, to interrogate this idea of the time
trade-off, we ran a subsequent regression with time spent on paid work
as the outcome variable. This regression includes gender and childcare
as explanatory variables, with the $(gender \times childcare)$
interaction term. Here, we found significant results and a clear visual
relationship of more time spent on childcare meant less time spent on
paid work. This upheld our expectations, as both women and men exhibited
this relationship.

## What are your key explanatory variables?

-   Gender
-   Time spent on childcare
-   Time spent on unpaid domestic work

## Research Questions

Is there a gender difference in how individuals allocate their time? How
does the trade-off between time spent at paid work versus time spent on
childcare differ by gender? Ultimately, the question of interest is how
do demographics and expected responsibilities affect time use.

# Questions: Data Wrangling and Transformation

## What data cleaning did you do?

-   Converted the time variables from minutes to hours for easy
    interpretation
-   Removed rows with NAs
-   Other than that, the diary entry data was well cleaned and coded by
    IPUMs! 🙂
-   We focus on 2010-2018 samples, since historical data prior to then
    is sporadic and requires time series methods.

## How did you wrangle the data?

We applied the following filters for the linear regression:

-   Married with kids: HHTYPE == 1
-   Child under 18: UNDER18 1 through 10
-   Post-college education: EDUC == 5 or 6
-   Full-time workers: FULLTIME == 1 (Only comparing men and women who
    work full-time)

Filtering was slightly less strict for Figures 1 through 4 (exploratory
visualizations).

## Are you deciding to exclude any observations? If so, why?

Yes, but this was mostly the data wrangling step. We removed any missing
values or NAs. For some visualizations (Figure 2) a few of the time use
variables were omitted to better depict explanatory variables, rather
than activities we’re less interested in: such as travel, media
consumption, physical activity, etc.

## Did you have to create any new variables from existing variables? If so, how and why?

Yes, when performing PCA on the time variables, we decided to combine a
few variables to increase interpretability. ACT_INHOME and ACT_OUTHOME
represented two types of leisure, so were combined into ACT_LEISURE.
ACT_PCARE and ACT_PHYSICAL, which represented personal care and physical
activity, were combined into ACT_SELFCARE since physical activity
(exercise) is also a form of personal/self care.

We also converted several variables into factors, to be used as
categorical variables in our visualizations and preliminary analysis.
For instance, SEX (formerly coded as 1 and 2) was re-coded into a binary
FEMALE categorical variable, 0 for men and 1 for female.

# Data Cleaning

This is the primary data cleaning and transforming done.

## Filtering for Visualizations 1-4

```{r}
filtered_data <- data %>%
  filter(
    HHTYPE == 1, # filter to married with kids
                 # isolating a group where childcare is relevant
                 # avoid comparing people who don’t face childcare constraints
    UNDER18 > 0, # filter to individuals with kids under age of 18
    AGE >= 22 & AGE <= 40, # between the ages of 22 and 40
    EDUC %in% c(5, 6), # college educated
    AGEYNGST >= 0 & AGEYNGST <= 18, # youngest child is between age 0 and 18
    SEX %in% c(1, 2), # male or female
    !is.na(AGE), 
    !is.na(EDUC)
  ) %>%
  mutate(
    sex_label = ifelse(SEX == 1, "Male", "Female")
  )
```

## Filtering for Visualization 5 (Regression Model)

```{r clean_transform}

# filtering and turning sex and full-time into binary variables
datareg <- filtered_data %>%
  
  filter(
    (EDUC == 5 | EDUC == 6), # filter to just college educated individuals
    FULLTIME == 1, # filter to just full-time workers
    WAGELM > 0, # remove non-earners
    ACT_CHCARE >= 0, # filter out missing or unclean data
    ACT_PCARE >= 0 # filter out missing or unclean data
  ) %>% 
  
  mutate(
    across(starts_with("ACT_"), ~ .x / 60), # convert all time use vars to hours
    female = ifelse(SEX == 2, 1, 0),
    female = factor(female),
    FULLTIME = factor(FULLTIME), 
    log_wage = log(WAGELM + 1) # add column for log wage in case it becomes interesting
  )
  
  
# filter to just 2010s data
data2010s <- datareg %>% 
  filter(SAMPLE > 2009)

```

# Codebook

`https://github.com/ccaitlingo/westcoasters-finalproject/blob/main/README.md`

# Summary Tables

```{r time_vars_summary}
# time variables
vars <- c("ACT_CHCARE", "ACT_CIVIC", "ACT_EDUCA", "ACT_INHOME", "ACT_MEDIA", "ACT_MISSING", "ACT_OUTHOME", "ACT_PCARE", "ACT_PHYSICAL", "ACT_TRAVEL", "ACT_UNDOM", "ACT_WORK")

means <- data2010s %>% summarise(across(all_of(vars), ~mean(.x, na.rm = TRUE))) %>% unlist()
sds   <- data2010s %>% summarise(across(all_of(vars), ~sd(.x, na.rm = TRUE)))   %>% unlist()

summary_stats <- tibble(
  variable = vars,
  mean = means,
  standard_deviation = sds
)

summary_stats
```

```{r categorical_vars_summary}

data2010s %>%
  group_by(female) %>%
  summarize(
    count = n(),
    percentage = (n() / nrow(data2010s)) * 100
  )
```

# Exploratory Data Analysis

## Motivation

### Figure 1: Unpaid Domestic Work by Sex

This visualization depicts the average hours per day that each sex (male
or female) spends performing unpaid domestic work, from 1975 to 2018.
Since our main outcome variable is time spent on **paid work**, it is
also important to complement that variable by considering time spent on
**unpaid work**, to get the full picture.

We can see that over time, the gap between male and female has shrunk.
However, a disparity still remains: even our most current data from 2018
shows that males on average spend 1.57 hours on unpaid domestic work,
while females spend 2.72 hours.

Note that there are fewer observations near 20

```{r Figure1}
# Unpaid Domestic Work by Sex (1930-2018)

# Average hours by year and sex
summary_data <- filtered_data %>%
  group_by(YEAR, sex_label) %>%
  mutate(ACT_UNDOM_HRS = ACT_UNDOM / 60) %>%
  summarise(avg_domestic_hours = mean(ACT_UNDOM_HRS, na.rm = TRUE), .groups = "drop")

# Line plot
ggplot(summary_data, aes(x = YEAR, y = avg_domestic_hours, color = sex_label)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "Unpaid Domestic Work by Sex (1975 - 2018)",
    x = "Year",
    y = "Average Hours per Day",
    color = "Sex"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("Male" = "steelblue", "Female" = "salmon"))

# Note for 2018
# 2.72 hrs for female
# 1.57 hrs for male
```

### Figure 2: Average Daily Time Use by Sex

This visualization breaks down the 24 hours in a day into the proportion
of time spent on various activities, averaged across the years
2010-2018. One of the time activities is our outcome variable, time
spent on **paid work**, while others include time spent performing
childcare, personal care, unpaid domestic work, and leisure.

```{r Figure2, message=FALSE, warning=FALSE}
# How time is split by gender on an average day!

# data wrangling
df <- filtered_data %>%
  filter(SAMPLE >= 2010 ) %>%  
  mutate(HHTYPE = as.numeric(as.factor(HHTYPE))) %>%
  mutate(
    Leisure = ACT_INHOME + ACT_OUTHOME
  ) %>% 
  select(-"ACT_CIVIC", -"ACT_EDUCA", -"ACT_MEDIA", -"ACT_MISSING",-"ACT_INHOME", -"ACT_OUTHOME", -"ACT_TRAVEL", -"ACT_PHYSICAL" )
# pivot! to have all activities listed under activity_category
df_long <- df %>%
  pivot_longer(
    cols = starts_with("ACT_") | starts_with("Leisure"),
    names_to = "activity_category",
    values_to = "minutes"
  )  %>% 
  mutate(
    # Clean up the activity names (optional)
    activity_category = gsub("ACT_", "", activity_category),
    activity_category = case_when(
      activity_category == "CHCARE" ~ "Childcare",
      activity_category == "PCARE"  ~ "Personal Care",
      activity_category == "UNDOM"  ~ "Unpaid Domestic Work",
      activity_category == "WORK"   ~ "Paid Work",
      TRUE ~ activity_category
    ),
    # Convert SEX codes to labels
    SEX = as.numeric(as.factor(SEX)), 
    SEX = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female"
    )
  )

# aggregate the activities!
df_sum <- df_long %>%
  group_by(SEX, activity_category) %>%
  summarize(minutes = mean(minutes, na.rm = TRUE))

sample_sizes <- df %>%
  mutate(SEX_label = ifelse(as.numeric(as.factor(SEX)) == 1, "Male", "Female")) %>%
  group_by(SEX_label) %>%
  summarize(n = n()) %>%
  # Create a combined label like "Male (n=500)"
  mutate(label_with_n = paste0(SEX_label, "\n(n = ", scales::comma(n), ")"))

#plot
ggplot(df_sum, aes(x = SEX, y = minutes, fill = activity_category)) +
  geom_bar(stat = "identity", position = "fill") +
  # Add the labels at the top
  geom_text(
    data = sample_sizes, 
    aes(x = SEX_label, y = 1.02, label = paste0("n = ", scales::comma(n))), 
    inherit.aes = FALSE, # This ignores the 'fill' requirement from the main ggplot
    vjust = 0,            # Aligns text above the point
  ) +
  labs(
    x = "Gender",
    y = "Proportion of Time Spent per Day",
    fill = "Activity",
    title = "Average Daily Time Use by Sex 2010-2018"
  ) +
  # Expand limits slightly so the text doesn't get cut off
  expand_limits(y = 1.1) + 
  theme_minimal()
```

### Figure 3: Clustering in 2D

This visualization uses hierarchical clustering to find groupings in
**PCA space**. We discovered that data points in clusters 2 and 3, which
represent the top right of the PCA space, are 69.8% female. This tells
us that females are more associated with having younger children, and
spending more time on unpaid domestic work and personal care!

**Interpreting Principal Components (PCS):** See the output summary
generated by the code, where the first four PCs are defined. Each PC is
a vector that represents a combination of factors. The most influential
factors that make up each PC are described as follows.

**PCA 1:** Time spent on childcare, age of the youngest child, and the
amount of children under 5

**PCA 2:** Time spent at work, time spent on self care, and time spent
doing unpaid domestic work

```{r Figure3, warning=FALSE, message=FALSE}

# Combine two leisure categories
filtered_data$ACT_LEISURE <- filtered_data$ACT_INHOME + filtered_data$ACT_OUTHOME
filtered_data$ACT_SELFCARE <- filtered_data$ACT_PCARE + filtered_data$ACT_PHYSICAL

# Select variables
vars <- c(
  "ACT_WORK", "ACT_CHCARE", "ACT_EDUCA",
  "ACT_LEISURE", "ACT_MEDIA", "ACT_TRAVEL", "ACT_SELFCARE",
  "ACT_UNDOM", "ACT_CIVIC", "AGEYNGST", "UNDER5", "UNDER18"
)

# Keep only selected variables + sex
df_vars <- filtered_data[, c(vars, "SEX")]

# Drop missing values
df_vars <- na.omit(df_vars)

# Sample data (original dataset too large)
set.seed(123)
df_sample <- df_vars[sample(nrow(df_vars), 5000), ]

# Create labels
df_sample$binary_var <- ifelse(df_sample$SEX == 1, "Male", "Female")

# Standardize the variables
df_scaled <- scale(df_sample[, vars])

# Run PCA
pca_res <- prcomp(df_scaled, center = TRUE, scale. = TRUE)

# Optional scree plot (tells us to use 4 PCs for optimal analysis)
eig_vals <- pca_res$sdev^2
var_explained <- eig_vals / sum(eig_vals)

plot(var_explained, type = "b",
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     main = "Scree Plot (PCA)",
     pch = 16)

# View the principal components
pca_res$rotation[, 1:4]

# Extract first 4 PCs
pca_scores <- as.data.frame(pca_res$x[, 1:4])

# Hierarchical clustering on PCA space
dist_mat <- dist(pca_scores)
hc <- hclust(dist_mat, method = "ward.D2")

# Cut into clusters
k <- 5
clusters <- cutree(hc, k = k)

# Attach cluster + sex
pca_scores$cluster <- factor(clusters)
pca_scores$binary_var <- df_sample$binary_var

# -------------------------
# Plot 1: PCA Space Colored by Sex
# -------------------------
plot(pca_scores$PC1, pca_scores$PC2,
     col = ifelse(pca_scores$binary_var == "Male", "blue", "red"),
     pch = 1,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "PCA Space Colored by Sex")

legend("topright",
       legend = c("Male", "Female"),
       col = c("blue", "red"),
       pch = 16,
       bty = "n")

# -------------------------
# Plot 2: Clusters in PCA Space
# -------------------------
cluster_colors <- c("#E69F00", "#56B4E9", "#009E73", "#CC79A7", "#F0E442")

plot(pca_scores$PC1, pca_scores$PC2,
     col = cluster_colors[pca_scores$cluster],
     pch = ifelse(pca_scores$binary_var == "Male", 1, 1),
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Clusters in PCA Space")

legend("topright",
       legend = c("Cluster 1","Cluster 2","Cluster 3","Cluster 4", "Cluster 5"),
       col = cluster_colors,
       pch = 16,
       bty = "n")

# -------------------------
# Table
# -------------------------
# Counts
tab <- table(pca_scores$cluster, pca_scores$binary_var)

# Row-wise percentages (within cluster)
pct <- round(prop.table(tab, margin = 1) * 100, 1)
pct

# Group clusters 2+3 (representing the top right of graph) and 1+4+5
grp <- ifelse(pca_scores$cluster %in% c(2,3), "C2+3", "C1+4+5")
pct_grp <- round(prop.table(table(grp, pca_scores$binary_var), 1) * 100, 1)
pct_grp
```

### Figure 4: Clustering in 3D

This visualization shows the same clustering in PCA space, but in three
dimensions instead of two. This allows us to visualize the third PC.

**PC3:** Time spent on leisure, time spent on travel

```{r Figure4, warning=FALSE, message=FALSE}
library(plotly)

# 3D PCA plot
plot_ly(
  pca_scores,
  x = ~PC1,
  y = ~PC2,
  z = ~PC3,
  color = ~as.factor(cluster),
  colors = cluster_colors,
  type = "scatter3d",
  mode = "markers"
)
```

## Preliminary Analysis

### Figure 5: Exploring the Research Question

In this visualization, we begin to observe the time trade-off outlined
in our research question and hypothesis - there appears to be an
opportunity cost associated with time spent at paid work versus time
spent on childcare. There's not an apparent difference between gender,
however there's a substantial split when time spent on childcare is less
than 1 hour a day. While this could be noise, maybe we could interrogate
this further in a regression. Otherwise, the gender split is minimal!
Both exhibit a clear inverse relationship between time spent on paid
work relative to time spent on childcare.

```{r Figure5, warning=FALSE, message=FALSE}

data2010s %>%
  mutate(chcare_bin = ntile(ACT_CHCARE, 10)) %>%
  group_by(chcare_bin, female) %>%
  summarise(
    avg_wage = mean(ACT_WORK, na.rm = TRUE),
    avg_chcare = mean(ACT_CHCARE)
  ) %>%
  ggplot(aes(x = avg_chcare, y = avg_wage, color = female)) +
  geom_line() +
  geom_point() +
    labs(
      title = "Time Spent on Paid Work vs. Time Spent on Childcare by Gender",
      subtitle = "In the 2010s for Individuals who are College-Educated and Married with Kids Under 18",
      x = "Average Hours of Childcare per Day",
      y = "Average Hours of Paid Work per Day"
    ) 

```
