---
title: "Final Project Prelim Data Analysis"
format: html
---

/Users/meredithhillier/Desktop/School/8. Spring 2026/MATH 244 - Intermediate Data Science/Final Project/westcoasters-finalproject


```{r load_packages, warnings=FALSE}
library(ipumsr)
library(tidyverse)
library(haven)
```

```{r load_data}
ddi <- read_ipums_ddi("data/ahtus_00001.xml")
data <- read_ipums_micro(ddi)
```

# Exploring proportion of women who get married vs men who get married over time

```{r}
marriedproportion <- data %>% 
  mutate(
    married = case_when(CIVSTAT == 04 ~ 0, TRUE ~ 1),
    SAMPLE = as.numeric(SAMPLE),
    SEX = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(SEX)) %>%
  group_by(SAMPLE, SEX) %>% 
  summarize(mean_married = mean(married), .groups = "drop")

marriedproportion
```

```{r}
ggplot(marriedproportion, aes(x = SAMPLE, y = mean_married, color = SEX)) + 
  geom_line() + labs(x = "Year", title = "Proportion of Individuals Married by Sex")
```


# Average time spent on child care for men vs women over time

```{r}
childcare_avg <- data %>% 
  mutate(
    SAMPLE = as.numeric(SAMPLE),
    SEX = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(SEX)) %>%
  group_by(SAMPLE, SEX) %>% 
  summarize(
    avg_childcare = mean(ACT_CHCARE, na.rm = TRUE),
    .groups = "drop"
  )

childcare_avg
```

```{r}
ggplot(childcare_avg, aes(x = SAMPLE, y = avg_childcare, color = SEX)) + 
  geom_line(size = 1.2) +
  labs(
    title = "Average Daily Child Care Time by Sex",
    x = "Year",
    y = "Minutes per Day",
    color = "Sex"
  ) 
```

# 4/13 Data Exploration

```{r}
data <- data %>% mutate(SEX2 = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female",
      TRUE ~ NA_character_
    ))
```


Childcare vs. wages, separately by sex
- Want a way to control for hours worked per day
- FULLTIME = 1 if full time work
- HHTYPE = 1 if married with child

```{r}
data %>%
  filter(!is.na(WAGELM), !is.na(ACT_CHCARE)) %>%
  filter(SAMPLE==2018, FULLTIME==1, HHTYPE==1) %>%
  ggplot(aes(x = ACT_CHCARE, y = WAGELM, color = SEX2)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship Between Childcare Time and Wages by Gender in 2018",
    x = "Minutes of Childcare per Day",
    y = "Monthly Wage"
  )
```


Distribution of wages by gender

```{r}
data %>%
  filter(!is.na(WAGELM)) %>%
  ggplot(aes(x = WAGELM, fill = SEX2)) +
  geom_density(alpha = 0.4) +
  labs(title = "Wage Distribution by Gender")
```
 
 Distribution of time spent on childcare by gender
 
```{r}
data %>%
  filter(!is.na(ACT_CHCARE)) %>%
  ggplot(aes(x = ACT_CHCARE, fill = SEX2)) +
  geom_density(alpha = 0.4) +
  labs(title = "Distribution of Childcare Time by Gender")
```

 
# REGRESSIONS

```{r cleaning}

# filtering and turning sex and full-time into binary variables
data <- data %>% 
  filter(HHTYPE == 1) %>% # filter to married with kids
                          # isolating a group where childcare is relevant
                          # avoid comparing people who don’t face childcare constraints
  filter(
    SEX %in% c(1, 2),
    FULLTIME %in% c(0, 1)
  ) %>% # filter out missing or dirty cases of sex and full time work
  
  filter(!is.na(AGE), !is.na(EDUC)) %>%
  mutate(
    female = ifelse(SEX == 2, 1, 0),
    female = factor(female),
    FULLTIME = factor(FULLTIME)
  ) %>%
  
  filter(FULLTIME == 1)

# filter to just 2018 data
data2018 <- data %>% filter(SAMPLE==2018, !is.na(WAGELM), !is.na(ACT_CHCARE))

```

```{r plots}

# plots
data2018 %>% select(WAGELM, ACT_CHCARE, WKHRS) %>% pairs

data2018 %>%
  ggplot(aes(x = ACT_CHCARE, y = WAGELM, color = SEX2)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    labs(
      title = "Relationship Between Childcare Time and Wages by Gender in 2018 for Individuals Who are Married with Kids",
      x = "Minutes of Childcare per Day",
      y = "Monthly Wage"
    ) 
```

Is childcare penalized more for women, controlling for employment intensity?

```{r}
baseline_model <- lm(WAGELM ~ ACT_CHCARE*female + AGE + EDUC, data2018)
summary(baseline_model)
```


```{r}
interaction_model <- lm(WAGELM ~ ACT_CHCARE*SEX*FULLTIME, data2018)
summary(interaction_model)
```

```{r}
interaction_model2 <- lm(WAGELM ~ ACT_CHCARE*SEX + ACT_CHCARE*FULLTIME, data2018)
summary(interaction_model2)
```

