---
title: "Final Project Prelim Data Analysis"
format: html
---

cd "/Users/meredithhillier/Desktop/School/8. Spring 2026/MATH 244 - Intermediate Data Science/Final Project/westcoasters-finalproject"


```{r load_packages, warnings=FALSE}
library(ipumsr)
library(tidyverse)
library(haven)
```

```{r load_data}
ddi <- read_ipums_ddi("data/ahtus_00001.xml")
data <- read_ipums_micro(ddi)
```

# Exploring proportion of women who get married vs men who get married over time

```{r}
marriedproportion <- data %>% 
  mutate(
    married = case_when(CIVSTAT == 04 ~ 0, TRUE ~ 1),
    SAMPLE = as.numeric(SAMPLE),
    SEX = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(SEX)) %>%
  group_by(SAMPLE, SEX) %>% 
  summarize(mean_married = mean(married), .groups = "drop")

marriedproportion
```

```{r}
ggplot(marriedproportion, aes(x = SAMPLE, y = mean_married, color = SEX)) + 
  geom_line() + labs(x = "Year", title = "Proportion of Individuals Married by Sex")
```


# Average time spent on child care for men vs women over time

```{r}
childcare_avg <- data %>% 
  mutate(
    SAMPLE = as.numeric(SAMPLE),
    SEX = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(SEX)) %>%
  group_by(SAMPLE, SEX) %>% 
  summarize(
    avg_childcare = mean(ACT_CHCARE, na.rm = TRUE),
    .groups = "drop"
  )

childcare_avg
```

```{r}
ggplot(childcare_avg, aes(x = SAMPLE, y = avg_childcare, color = SEX)) + 
  geom_line(size = 1.2) +
  labs(
    title = "Average Daily Child Care Time by Sex",
    x = "Year",
    y = "Minutes per Day",
    color = "Sex"
  ) 
```

# 4/13 Data Exploration

```{r}
data <- data %>% mutate(SEX2 = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female",
      TRUE ~ NA_character_
    ))
```


Childcare vs. wages, separately by sex
- Want a way to control for hours worked per day
- FULLTIME = 1 if full time work
- HHTYPE = 1 if married with child

```{r}
data %>%
  filter(!is.na(WAGELM), !is.na(ACT_CHCARE)) %>%
  filter(SAMPLE==c(2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018), FULLTIME==1, HHTYPE==1) %>%
  ggplot(aes(x = ACT_CHCARE, y = WAGELM, color = SEX2)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship Between Childcare Time and Wages by Gender in the 2010s",
    x = "Minutes of Childcare per Day",
    y = "Monthly Wage"
  )
```


Distribution of wages by gender

```{r}
data %>%
  filter(!is.na(WAGELM)) %>%
  ggplot(aes(x = WAGELM, fill = SEX2)) +
  geom_density(alpha = 0.4) +
  labs(title = "Wage Distribution by Gender")
```
 
 Distribution of time spent on childcare by gender
 
```{r}
data %>%
  filter(!is.na(ACT_CHCARE)) %>%
  ggplot(aes(x = ACT_CHCARE, fill = SEX2)) +
  geom_density(alpha = 0.4) +
  labs(title = "Distribution of Childcare Time by Gender")
```

 
# REGRESSIONS

```{r clean_transform}

# filtering and turning sex and full-time into binary variables
data <- data %>%
  
  filter(
    HHTYPE == 1, # filter to married with kids
                 # isolating a group where childcare is relevant
                 # avoid comparing people who don’t face childcare constraints
    SEX %in% c(1, 2), # filter out missing or dirty cases of sex
    # FULLTIME %in% c(0, 1), # filter out missing or dirty cases of full time work
    UNDER18 > 0, # filter to individuals with kids under age of 18
    !is.na(AGE), 
    !is.na(EDUC),
    FULLTIME == 1, # filter to just full-time workers
    (EDUC == 5 | EDUC == 6), # filter to just college educated individuals
    WAGELM > 0, # remove non-earners
    ACT_CHCARE >= 0,
    ACT_PCARE >= 0
  ) %>% 
  
  mutate(
    across(starts_with("ACT_"), ~ .x / 60), # convert all time use vars to hours
    female = ifelse(SEX == 2, 1, 0),
    female = factor(female),
    FULLTIME = factor(FULLTIME), 
    log_wage = log(WAGELM + 1) # add column for log wage in case it becomes interesting
  )
  
  
# filter to just 2010s data
data2010s <- data %>% 
  filter(SAMPLE > 2009)

```

```{r plots, include=FALSE}

# plots
#data2010s %>% select(WAGELM, ACT_CHCARE, WKHRS) %>% pairs

data2010s %>%
  ggplot(aes(x = ACT_CHCARE, y = WAGELM, color = SEX2)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    labs(
      title = "Relationship Between Childcare Time and Wages by Gender in 2010s for Individuals Who are Married with Kids",
      x = "Hours of Childcare per Day",
      y = "Monthly Wage"
    ) 

data2010s %>%
  mutate(chcare_bin = ntile(ACT_CHCARE, 5)) %>%
  group_by(chcare_bin, female) %>%
  summarise(
    avg_wage = mean(WAGELM, na.rm = TRUE),
    avg_chcare = mean(ACT_CHCARE)
  ) %>%
  ggplot(aes(x = avg_chcare, y = avg_wage, color = female)) +
  geom_line() +
  geom_point() +
    labs(
      title = "Relationship Between Childcare Time and Wages by Gender in 2010s for Individuals Who are Married with Kids",
      x = "Average Hours of Childcare per Day",
      y = "Average Monthly Wage"
    ) 

data2010s %>%
  mutate(chcare_cat = cut(ACT_CHCARE, breaks = 5)) %>%
  ggplot(aes(x = chcare_cat, y = WAGELM, fill = female)) +
  geom_boxplot() +
    labs(
      title = "Relationship Between Childcare Time and Wages by Gender in 2010s for Individuals Who are Married with Kids",
      x = "Average Hours of Childcare per Day",
      y = "Average Monthly Wage"
    ) 
```

```{r plots2, include=FALSE}

data2010s %>%
  ggplot(aes(x = ACT_CHCARE, y = ACT_PCARE, color = SEX2)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    labs(
      title = "Relationship Between Childcare Time and Personal Care by Gender in 2010s for Individuals Who are Married with Kids",
      x = "Hours of Childcare per Day",
      y = "Hours of Personal Care per Day"
    ) 

data2010s %>%
  mutate(chcare_bin = ntile(ACT_CHCARE, 5)) %>%
  group_by(chcare_bin, female) %>%
  summarise(
    avg_wage = mean(ACT_PCARE, na.rm = TRUE),
    avg_chcare = mean(ACT_CHCARE)
  ) %>%
  ggplot(aes(x = avg_chcare, y = avg_wage, color = female)) +
  geom_line() +
  geom_point() +
    labs(
      title = "Relationship Between Childcare Time and Wages by Gender in 2010s for Individuals Who are Married with Kids",
      x = "Average Hours of Childcare per Day",
      y = "Average Monthly Wage"
    ) 

data2010s %>%
  mutate(chcare_cat = cut(ACT_CHCARE, breaks = 5)) %>%
  ggplot(aes(x = chcare_cat, y = ACT_PCARE, fill = female)) +
  geom_boxplot() +
    labs(
      title = "Relationship Between Childcare Time and Wages by Gender in 2010s for Individuals Who are Married with Kids",
      x = "Average Hours of Childcare per Day",
      y = "Average Monthly Wage"
    ) 
```


```{r plots}

# plots
#data2010s %>% select(WAGELM, ACT_CHCARE, WKHRS) %>% pairs

data2010s %>%
  ggplot(aes(x = ACT_CHCARE, y = ACT_WORK, color = SEX2)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    labs(
      title = "Time Spent on Paid Work vs. Time Spent on Childcare by Gender",
      subtitle = "In the 2010s for Individuals who are College-Educated and Married with Kids Under 18",
      x = "Hours of Childcare per Day",
      y = "Hours of Paid Work per Day"
    ) 

data2010s %>%
  mutate(chcare_bin = ntile(ACT_CHCARE, 5)) %>%
  group_by(chcare_bin, female) %>%
  summarise(
    avg_wage = mean(ACT_WORK, na.rm = TRUE),
    avg_chcare = mean(ACT_CHCARE)
  ) %>%
  ggplot(aes(x = avg_chcare, y = avg_wage, color = female)) +
  geom_line() +
  geom_point() +
    labs(
      title = "Time Spent on Paid Work vs. Time Spent on Childcare by Gender",
      subtitle = "In the 2010s for Individuals who are College-Educated and Married with Kids Under 18",
      x = "Average Hours of Childcare per Day",
      y = "Average Hours of Paid Work per Day"
    ) 

data2010s %>%
  mutate(chcare_cat = cut(ACT_CHCARE, breaks = 5)) %>%
  ggplot(aes(x = chcare_cat, y = ACT_WORK, fill = female)) +
  geom_boxplot() +
    labs(
      title = "Time Spent on Paid Work vs. Time Spent on Childcare by Gender",
      subtitle = "In the 2010s for Individuals who are College-Educated and Married with Kids Under 18",
      x = "Average Hours of Childcare per Day",
      y = "Average Hours of Paid Work per Day"
    ) 

```


Base regression

```{r}
base_model <- lm(ACT_WORK ~ ACT_CHCARE*female, data2010s)
summary(base_model)
```

```{r}
base_model2 <- lm(ACT_WORK ~ ACT_CHCARE*female + ACT_PCARE + ACT_EDUCA + ACT_UNDOM + ACT_CIVIC + ACT_OUTHOME + ACT_PHYSICAL + ACT_INHOME + ACT_MEDIA + ACT_TRAVEL + ACT_MISSING, data2010s)
summary(base_model2)
```




Base regression with controls

```{r}
base_model_controls <- lm(ACT_WORK ~ ACT_CHCARE*female + AGE + EDUC + WKHRS + UNDER18 + AGEYNGST + URBAN, data2010s)
summary(base_model_controls)
```




Is childcare penalized more for women, controlling for employment intensity?

```{r}
baseline_model <- lm(WAGELM ~ ACT_CHCARE*female + ACT_UNDOM*female + AGE + EDUC, data2010s)
summary(baseline_model)
```


```{r}
interaction_model <- lm(WAGELM ~ ACT_CHCARE*SEX*FULLTIME, data2010s)
summary(interaction_model)
```

```{r}
interaction_model2 <- lm(WAGELM ~ ACT_CHCARE*SEX + ACT_CHCARE*FULLTIME, data2010s)
summary(interaction_model2)
```

